\documentclass[12pt,openany]{book}

\usepackage[margin=0.9in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage[T1]{fontenc}

\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]

\usepackage[hidelinks]{hyperref}
\hypersetup{
  pdftitle={Linear, Geometric, and Structural Transformations: An Axiomatic and Coordinate-Free Perspective},
  pdfauthor={Jos\'e Portillo},
  pdfsubject={Linear algebra, geometry, tensors, and logarithmic structure},
  pdfkeywords={linear transformations, inner products, tensors, tensor product, logarithms}
}

\begin{document}

\pagestyle{plain}

\frontmatter

\begin{titlepage}
\thispagestyle{empty}

\vspace*{2.2cm}

{\Large\bfseries
Linear, Geometric, and Structural Transformations\par}
\vspace{0.4cm}

\vspace{1.4cm}
{\Large Jos\'e Portillo\par}

\vspace{1.8cm}
\rule{\textwidth}{0.6pt}
\vspace{0.35cm}

{\small
\textbf{Release date:} February 15, 2026\par
\textbf{Version:} 0.4\par
}

\vfill

\end{titlepage}


\thispagestyle{empty}
\null\vfill
\begin{center}
To the memory of\par
Mercelena Portillo Holskin,\par
Eldon C. Hall,\par
Hugh Blair-Smith\par
\vspace{1cm}
To my family
\end{center}
\vfill\null
\clearpage

\tableofcontents

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Every algorithm, however sophisticated, reduces to a finite sequence of
elementary operations: additions, multiplications, comparisons.
Every physical system that performs computation --- whether built from vacuum
tubes, transistors, optical waveguides, or any substrate yet to be conceived
--- implements those same operations through different physical means.
What persists (underlies) all such changes of medium and representation is
a layer of mathematical structure.
This monograph is concerned with that layer, which remains invariant.

\medskip

The question that organizes the exposition is therefore \emph{what} is being computed --- and what
structural properties that object possesses independently of the coordinate
system, basis, or hardware in which it happens to be expressed.
This monograph began as an attempt to understand what remains invariant when
formulas, coordinates, and representations are removed.

\medskip

A recurring distinction will be made between an axiomatic theory and a
concrete model of that theory.
In a refined axiomatic treatment of Euclidean geometry, objects such as
\textit{point}, \textit{line}, and \textit{plane} are taken as primitive
(undefined) terms, and the axioms specify only the relations they must satisfy.
A Cartesian coordinate system does not redefine these primitives; rather, it
provides a \emph{model} in which points are represented by tuples
in~$\mathbb{R}^n$, lines by solution sets of linear equations, and incidence
becomes set membership~\cite{iribarren2}.
As stated in~\cite{wilder}, this is \textit{the assignment of meanings}.
This perspective helps place coordinate formulas as representations of
structure, and not as the structure itself.
The same distinction runs through every chapter of this text.

\medskip

Vector spaces are introduced as abstract sets equipped with addition and scalar
multiplication.
Linear transformations are defined as the maps that preserve this structure.

\medskip

By introducing bilinear forms and, ultimately, inner products, vectors are
allowed to interact in a way that produces scalars.
From this single interaction arise norms, orthogonality, distances, and
classical geometric objects such as balls, spheres, and ellipsoids.

\medskip

Many constructions encountered earlier --- bilinear forms, inner products,
matrix multiplication --- share a common feature: they are linear in each
argument separately.
This observation leads naturally to multilinear maps and to tensors.

\medskip

Tensors are treated as abstract multilinear objects defined without
coordinates.
The tensor product is introduced through its universal property, explaining
how multilinear problems are canonically reduced to linear ones.
Index notation, arrays, and computational implementations are presented
explicitly as representations.

\medskip

The text is organized around four successive structural themes: linear
structure, geometric and topological structure, multilinear structure, and
logarithmic structure.
Each theme is a refinement of the same organizing idea: structure is revealed
through invariance under transformations.
The logarithmic theme, in particular, illustrates that this principle is not
confined to the linear setting --- the logarithm is a global isomorphism
between multiplicative and additive structures, and as such it belongs to the
same invariant framework as vectors and tensors.

\medskip

For the reader, in order to understand the subjects and use these notes as a
starting point for further research, prior familiarity is assumed with basic
set theory~\cite{wilder}\cite{cohen}\cite{apostol}, relations, functions
\cite{apostol}, $n$-tuples~\cite{iribarren}\cite{cohen}, and basic
algebra~\cite{iribarren}.
Basic knowledge of calculus~\cite{apostol} is also assumed.
The reports~\cite{kolecki} and~\cite{kolecki2} cover the foundational material
on tensors.
Other introductory material is scattered throughout the bibliography.

\medskip

The topics gathered in this monograph belong to the established literature on
linear algebra and abstract algebra.
What motivated its writing was therefore the perception that a single organizing idea ---
the preservation of structure under transformation --- recurs persistently
across both pure and applied science.
To gather that unity into a single volume is the purpose of this work.
Throughout the text, the word \textit{invariant} is used in a broad sense:
that which remains unchanged when coordinates, representations, or physical
substrates are varied.
The reader acquainted with algebra will recognize that the term also carries
more precise technical meanings --- invariant subspaces, similarity invariants,
topological invariants --- which should not be confused with this structural
and transversal usage.

\medskip

An AI language model was used for editing, \LaTeX{} and stylistic refinement.

\medskip

This monograph presents the abstract structure that underlies every algorithm
and autonomous computing system, and therefore remains invariant under any
change in the state of technology.
There is a more physical and hardware-oriented counterpart, centered on a
1960s computer --- the Apollo Guidance Computer --- which is more specific to a
moment in the history of the art.
For the full context, see:

\medskip

{\small
\url{https://jportillo34.github.io/ApolloGuidanceComputer/}
}

\clearpage

\makeatletter
\@mainmattertrue
\makeatother
\pagenumbering{arabic}

\include{Structures_I}
\include{Structures_II}
\include{Structures_III}
\include{Structures_IV}

\backmatter
\clearpage
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}

\begin{thebibliography}{99}

\bibitem{iribarren}
Iribarren Ignacio L., \textit{Álgebra Lineal}, Editorial Equinoccio, 2009.

\bibitem{halmos}
Halmos, Paul R., \textit{Finite-Dimensional Vector Spaces}, Springer-Verlag, 1987.

\bibitem{iribarren2}
Iribarren, Ignacio L., \textit{Topología de Espacios Métricos}, Editorial Limusa-Wiley, 1973.

\bibitem{cohen}
Cohen, Leon and Ehrlich, Gertrude, \textit{The Structure of the Real Number System}, D. Van Nostrand Company, 1963.

\bibitem{wilder}
Wilder, Raymond L., \textit{Introduction to The Foundations of Mathematics}, Dover Publications, Inc., 2012.

\bibitem{greub}
Greub, Werner, \textit{Linear Algebra}, Springer-Verlag, 1975.

\bibitem{oxford}
Lukas, Andre, \textit{The Oxford Linear Algebra for Scientists}, Oxford University Press, 2022.

\bibitem{apostol}
Apostol, Tom M., \textit{Calculus Volume I}, John Wiley \& Sons Inc., 1967.

\bibitem{knuth}
Knuth, Donald E., \textit{The Art of Computer Programming Volume I: Fundamental Algorithms}, Addison-Wesley, 1997.

\bibitem{knuth2}
Knuth, Donald E., \textit{The Art of Computer Programming Volume II: Seminumerical Algorithms}, Addison-Wesley, 1981.

\bibitem{knuth3}
Knuth, Donald E., \textit{The Art of Computer Programming Volume I, Fascicle 1: MMIX—A RISC Computer for the New Millennium}, Addison-Wesley, 2005.

\bibitem{feynman}
Feynman Richard P., Leighton Robert, Sands Matthew, \textit{The Feynman Lectures on Physics, Volume II}, California Institute of Technology, 2010.

\bibitem{kolecki}
Kolecki, Joseph C., \textit{An Introduction to Tensors for Students of Physics and Engineering}, NASA/TM-2002-211716, 2002.

\bibitem{kolecki2}
Kolecki, Joseph C., \textit{Foundations of Tensor Analysis for Students of Physics and Engineering With an Introduction
to the Theory of Relativity}, NASA/TP—2005-213115, 2005.

\bibitem{kung}
Kung, H. T., Liserson, Charles L., \textit{Introduction to VLSI Systems - Algorithms for VLSI Processor Arrays}, ed. C. Mead and L. Conway, Addison-Wesley, Reading, MA, 1980.

\bibitem{Xu}
Shaofu Xu, Jing Wang, Haowen Shu, Zhike Zhang, Sicheng Yi, Bowen Bai, XingjunWang, Jianguo Liu, Weiwen Zou, \textit{Optical coherent dot-product chip for sophisticated deep learning regression}, Nature - Light: Science \& Applications, 2021.

\end{thebibliography}

\end{document}
